---
title: "MitoDiveR: A User Manual for Comparative Mitochondrial Genomics"
author: "Michael W. Sandel"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{MitoDiveR: A User Manual for Comparative Mitochondrial Genomics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

# Overview

**MitoDiveR** is a toolkit for comparative mitochondrial genomics in R. It
covers the full arc of a mitogenome study:

1. **Fetch** annotated mitogenomes and gene features from NCBI.
2. **Discover** open reading frames (ORFs) — canonical genes and the
   noncanonical / mitochondrial-derived-peptide (MDP) ORFs that hide in
   alternate frames and on the light strand.
3. **Compare** ORFeomes across genomes and across whole clades to find
   conserved coding potential.
4. **Analyze** sequence variants (SNPs) for their protein, ORF, and RNA-structure
   consequences.
5. **Quantify** divergence with codon usage and dN/dS.
6. **Search** for MDP homologs and characterize peptides with structure
   (ESMFold) and topology (DeepTMHMM) predictions.
7. **Visualize** an entire six-frame ORFeome as a circos plot.

This manual walks through each stage. Most examples that touch the network
(NCBI) or external tools (ViennaRNA `RNAfold`, the ESMFold REST API, the
`biolib`/DeepTMHMM CLI) are shown but **not executed** here, so they are marked
`eval = FALSE`. A small self-contained example at the end runs end-to-end with
no network access.

```{r install, eval = FALSE}
# From GitHub (development version)
# install.packages("remotes")
remotes::install_github("evozoa/MitoDiveR")

library(MitoDiveR)
```

---

# The one rule you must know: `genetic_code` is always explicit

Mitochondrial genomes do **not** use the standard genetic code. Every
ORF/codon/variant function in MitoDiveR therefore **requires** you to state the
code explicitly — there is no default, on purpose. Omitting it raises an
informative error rather than silently guessing.

The two codes you will use:

| Code     | Name                             | Use for                                                                 |
|----------|----------------------------------|-------------------------------------------------------------------------|
| `"SGC1"` | Vertebrate mitochondrial         | The 13 canonical mtDNA protein genes (TGA = Trp; AGA/AGG = stop)         |
| `"SGC0"` | Standard                         | MDPs and other noncanonical ORFs translated by the cytoplasmic ribosome |

> **Rule of thumb.** Canonical mito-proteins → `"SGC1"`. MDPs / antisense /
> alt-frame ORFs that would be translated in the cytoplasm → `"SGC0"`. When in
> doubt, scan **both** and compare. Passing the wrong code is the single most
> common way to get biologically wrong results, which is exactly why the
> argument has no default.

(These are `Biostrings` genetic-code identifiers; see
`Biostrings::GENETIC_CODE_TABLE`.)

---

# 1. Fetching sequences and annotations

`fetch_sequences()` resolves input from NCBI accessions, a FASTA file, or raw
FASTA text into a `Biostrings::DNAStringSet`.

```{r fetch-seq, eval = FALSE}
# Human reference mitogenome (rCRS) plus chimpanzee, from NCBI
seqs <- fetch_sequences(accessions = c("NC_012920.1", "NC_001643.1"))

# ...or from local FASTA
seqs <- fetch_sequences(fasta_file = "my_mitogenomes.fasta")
```

`fetch_mito_genbank()` additionally parses the GenBank **feature table**, which
you need for gene-aware plots and conserved-window analyses.

```{r fetch-gb, eval = FALSE}
gb       <- fetch_mito_genbank("NC_012920.1")
features <- gb[["NC_012920.1"]]$features   # data.frame of CDS / rRNA / tRNA / D-loop
```

All ORF/SNP functions also accept accessions, FASTA paths, or raw sequence text
directly (via the internal input resolver), so you can usually skip the explicit
fetch step.

---

# 2. ORF discovery

## `find_orfs()` — a single sequence

```{r find-orfs, eval = FALSE}
orfs <- find_orfs(
  "NC_012920.1",
  genetic_code   = "SGC1",   # canonical vertebrate-mito scan
  start_codons   = "ATG",
  min_orf_length = 100L,
  both_strands   = TRUE,
  circular       = TRUE      # mtDNA is circular: ORFs may wrap the origin
)
```

## `scan_orfs()` — many sequences at once

`scan_orfs()` takes a `DNAStringSet` (or named character vector) and returns one
tidy table with a `seq_id` column. Its default `start_codons` include the
alternative mitochondrial starts (`GTG`, `ATT`, `ATC`, `ATA`).

```{r scan-orfs, eval = FALSE}
orf_table <- scan_orfs(
  seqs,
  genetic_code   = "SGC1",
  min_orf_length = 150L
)
```

### Nested-ORF inflation — and `collapse_nested_orfs()`

`find_orfs()`/`scan_orfs()` report **every** ATG→stop as its own ORF, so one
long stop-free stretch with many internal starts yields many entries that share
a single stop. Raw per-frame ORF *counts* are therefore not a clean metric.
Collapse to the longest ORF per stop before counting or plotting:

```{r collapse, eval = FALSE}
loci <- collapse_nested_orfs(orf_table)   # keeps the longest ORF per (frame, stop)
# e.g. human rCRS: ~505 raw ORFs -> 26 distinct loci
```

> **Minus-strand coordinates.** The table enforces genomic `start <= end`, so on
> the `-` strand the **stop** is in `start` and the **ATG** is in `end`. This is
> the genomic convention, not a bug.

## `scan_orfeome()` — sweep across parameters

To explore sensitivity to code, length cutoff, and start-codon set in one call:

```{r scan-orfeome, eval = FALSE}
sweep <- scan_orfeome(
  seqs,
  genetic_codes     = c("SGC1", "SGC0"),
  min_orf_lengths   = c(45L, 100L),
  start_codon_sets  = list("ATG", c("ATG", "GTG", "ATT"))
)
# Output is tagged with genetic_code / min_orf_length / start_codon_set columns
```

---

# 3. Conserved ORFs across genomes and clades

`find_conserved_orfs()` clusters ORFs from a `scan_orfs()` table by amino-acid
identity, surfacing coding potential shared across genomes.

```{r conserved, eval = FALSE}
conserved <- find_conserved_orfs(
  orf_table,
  min_identity  = 50,    # percent aa identity to cluster
  min_sequences = 0.5    # present in >= 50% of genomes
)
```

`find_clade_conserved_orfs()` does the whole pipeline in one call — NCBI search
for RefSeq mitogenomes in a taxon, fetch, scan, and cluster:

```{r clade, eval = FALSE}
res <- find_clade_conserved_orfs(
  taxon         = "Sebastes",   # rockfishes
  genetic_codes = "SGC1",
  max_genomes   = 20,
  refseq_only   = TRUE
)
res$conserved   # conserved-ORF clusters across the clade
```

To suppress spurious links from short high-identity motifs, add a coverage
gate: `find_conserved_orfs(orfs, min_coverage = 0.5)` requires each alignment
to span a minimum fraction of the shorter protein. (This subsumes the former
`cluster_conserved_orfs()`, now a deprecated shim — see `?find_conserved_orfs`.)

Related helpers: `find_conserved_windows()` (conserved nucleotide windows),
`compare_clade_orfeomes()`, and `compare_mitogenomes()` (pairwise mutation
table with per-SNP mRNA/MDP impact).

---

# 4. Variant (SNP) analysis

Two complementary entry points:

## `analyze_snp()` — variant-first

Give it a single substitution and it reports the consequences for noncanonical
ORFs (and, optionally, RNA structure).

```{r analyze-snp, eval = FALSE}
res <- analyze_snp(
  "m.1382A>C",                      # also accepts "A1382C" or list(pos, ref, alt)
  reference     = "NC_012920.1",    # defaults to human rCRS
  genetic_codes = "SGC0",           # MDP scan uses the standard code
  rna_structure = TRUE,             # opt-in: folds ref vs mutant window (needs RNAfold)
  rna_flank     = 100
)
res$summary   # per-code table: disrupted / novel / lost / gained ORFs
res$rna       # delta_mfe, base-pair distance, longest-stem change, structures
```

## `characterize_transcript_snps()` — transcript-first

Enumerate the SNPs observed across a transcript/region and characterize each.
For human, variants come from **MITOMAP**; for other taxa, from NCBI sampling.

```{r characterize, eval = FALSE}
mots_c <- characterize_transcript_snps(
  region        = "MOTS-c",          # preset, or list(start, end, strand)
  taxon         = "Homo sapiens",
  genetic_codes = "SGC0",
  rna_structure = TRUE
)
mots_c$variants   # pos/ref/alt/freq/consequence/n_disrupted/delta_mfe/...
```

Protein-level scoring of an amino-acid change (BLOSUM62, Chou–Fasman,
Kyte–Doolittle) is available via `analyze_protein_impacts()`.

---

# 5. Divergence: codon usage and dN/dS

```{r divergence, eval = FALSE}
usage <- calc_codon_usage(orf_table, genetic_code = "SGC1")

rates <- calc_dnds(orf_table, genetic_code = "SGC1")   # Nei-Gojobori + Jukes-Cantor

pos_rates <- calc_codon_position_rates(orf_table, genetic_code = "SGC1")
```

---

# 6. Mitochondrial-derived peptides (MDPs)

```{r mdp, eval = FALSE}
# Reference MDP sequences (humanin, MOTS-c, SHLP1-6, ...)
peptides <- mdp_sequences()

# Score similarity of a candidate ORF to known MDPs
score_mdp_similarity(candidate_aa = "MRWQEMGYIFYPRKLR", peptides = peptides)

# Search for homologs of an MDP across sequences
hits <- search_mdp_homologs(query = peptides["MOTS-c"], subjects = seqs)
```

> **Heads-up.** `mdp_sequences()` carries known errors for SHLP1–6; for those,
> prefer canonical UniProt sequences. SHLP3 in particular exists in both the
> human-derived `MLGY...` form and the ancestral `MQKVR...` form.

---

# 7. Structure and topology predictions

These call external services and are inherently optional / network-bound.

```{r struct, eval = FALSE}
# ESMFold structure + per-residue confidence (REST API)
pdb   <- predict_esmfold("MAPRGFSCLLLLTSEIDLPVKRRA")
plddt <- esmfold_plddt(pdb)

# Transmembrane topology via DeepTMHMM (biolib CLI)
topo    <- predict_tmhmm("MAPRGFSCLLLLTSEIDLPVKRRA")
domains <- tmhmm_domains(topo)
```

---

# 8. RNA secondary structure

Wraps ViennaRNA `RNAfold` (must be on `PATH`).

```{r rna, eval = FALSE}
fold <- fold_rna("GGGAAACCCUUUGGGAAACCC")
fold$mfe          # minimum free energy
fold$structure    # dot-bracket

pairs <- pair_partners(fold$structure)   # base-pair partner vector
longest_stem(fold$structure)             # longest contiguous helix
```

---

# 9. Visualizing the ORFeome: circos

`plot_orfeome_circos()` draws all six reading frames as concentric rings around
a central ring of nucleotide coordinates and canonical gene features. Pass the
GenBank features for gene labels; nested ORFs are collapsed by default so the
rings show distinct loci.

```{r circos, eval = FALSE}
gb    <- fetch_mito_genbank("NC_012920.1")
genes <- gb[["NC_012920.1"]]$features

plot_orfeome_circos(
  orfs          = orf_table[orf_table$seq_id == "NC_012920.1", ],
  genome_length = 16569,
  genes         = genes,
  seq_id        = "NC_012920.1",
  title         = "Human mtDNA ORFeome",
  file          = "human_orfeome_circos.png"   # omit to draw to the active device
)
```

Requires the `circlize` package (in `Suggests`).

---

# 10. Saving results

`save_results()` writes analysis tables/objects to disk in a consistent layout
so workflows are reproducible.

```{r save, eval = FALSE}
save_results(orf_table, file = "orfs.csv")
```

---

# A self-contained example

This runs without any network access or external tools — a tiny synthetic
sequence with one obvious forward ORF.

```{r selfcontained}
library(MitoDiveR)

# ATG ... (sense codons) ... TAA, padded so the ORF clears the length filter
seq <- paste0(
  "ATG",
  paste(rep("AAACTTGGCATCGTA", 4), collapse = ""),  # filler sense codons
  "TAA"
)

orfs <- find_orfs(
  seq,
  genetic_code   = "SGC0",   # standard code for this toy example
  start_codons   = "ATG",
  min_orf_length = 30L,
  both_strands   = FALSE,
  circular       = FALSE
)

orfs[, c("start", "end", "strand", "length_nt")]
```

---

# Where to go next

- Per-function help: `?find_orfs`, `?analyze_snp`, `?find_clade_conserved_orfs`, …
- Worked case studies backing the *Geromedicine* manuscript live under
  `analyses/mitogenome_comparison/` in the source repository (human, primate,
  cyprinid, killifish, rockfish, rodent, sculpin, urchin, xenopus, midichloria,
  and the vertebrate 16S frameshift).
- Report issues at <https://github.com/evozoa/MitoDiveR/issues>.

```{r sessioninfo}
sessionInfo()
```
